---
title: "Assignment 5: Data Visualization"
author: "Logan Dye"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Oct 14th @ 5:00pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse, 
lubridate, & cowplot packages. Upload the NTL-LTER processed data files for  
nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the
[`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
#getwd()
library(tidyverse)
library(lubridate)
library(cowplot)

PeterPaulChemNut <- 
  read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)
LitterData <-
  read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

#2 
str(LitterData)
LitterData$collectDate <- as.Date(LitterData$collectDate, format = "%Y-%m-%d")

str(PeterPaulChemNut)
PeterPaulChemNut$year4 <- as.Date(PeterPaulChemNut$sampledate, format = "%Y-%m-%d")

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "red"), 
        legend.position = "right")
theme_set(mytheme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best 
practices for data visualization. Ensure your theme, color palettes, axes, 
and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with 
separate aesthetics for Peter and Paul lakes. Add a line of best fit and 
color it black. Adjust your axes to hide extreme values (hint: change the 
limits using `xlim()` and/or `ylim()`).

```{r}
#4

PhosphorusByPhosphate <- 
  ggplot(PeterPaulChemNut, aes(x = po4, y = tp_ug)) +
  geom_point(aes(color = lakename)) +
  geom_smooth(method='lm', color = "black") + 
  xlim(0,75)+
  ylim(0, 175)
print(PhosphorusByPhosphate)


```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and 
(c) TN, with month as the x axis and lake as a color aesthetic. Then, create 
a cowplot that combines the three graphs. Make sure that only one legend is 
present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;
see <https://r-lang.com/month-abb-in-r-with-example>

```{r}
#5
PeterPaulChemNut$month <- factor(PeterPaulChemNut$month, levels=c(1:12))

TempBPlot <-
  ggplot(PeterPaulChemNut, aes(x = month, y = temperature_C)) + 
  geom_boxplot(aes(color = lakename)) +
  scale_x_discrete(label = month.abb, drop = FALSE) + 
  ylab("Temp_C") + 
  theme(legend.position = "none")
print(TempBPlot)

TPBPlot <-
  ggplot(PeterPaulChemNut, aes(x = month, y = tp_ug)) + 
  geom_boxplot(aes(color = lakename)) + 
  scale_x_discrete(label = month.abb, drop = FALSE) +
  theme(legend.position = "none")
print(TPBPlot)

TNBPlot <-
  ggplot(PeterPaulChemNut, aes(x = month, y = tn_ug)) + 
  geom_boxplot(aes(color = lakename)) +
  scale_x_discrete(label = month.abb, drop = FALSE) +
  theme(legend.position = "bottom")
print(TNBPlot)

plot_grid(TempBPlot, TPBPlot, TNBPlot, nrow = 3, align = "v", axis = c("l"))  

```

Question: What do you observe about the variables of interest over seasons 
and between lakes? 

> Answer: The temperatures between the two lakes move in relative synchronicity. Summer months have larger medians. This makes sense condidering their geographic relation to eachother and the fact that summer months are hotter. For TP, Peter lake consistantly has higher values than Paul Lake. TP for Peter lake consistantly rises from May through September, while Paul lake stays at relativly the same level. For TN, the medians are similar with Peter lake being slightly higher, but not by much. The difference between the lakes reveals itself in the high value outliers. Peter lake has upper level outliers that are of much greater value than Paul lake in every month. There is not a dramatic difference in TN seasonally. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the 
"Needles" functional group. Plot the dry mass of needle litter by date and 
separate by NLCD class with a color aesthetic. (no need to adjust the name 
of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into 
three facets rather than separated by color. 

```{r}
#6
LitterDataSpreadNeedles <- pivot_wider(LitterData, names_from = functionalGroup, values_from = dryMass)

NeedleMassPlot <- ggplot(LitterDataSpreadNeedles) +
  geom_point(aes(x = collectDate, y = Needles, color = nlcdClass))
print(NeedleMassPlot)

#7
NeedleMassFacetPlot <- ggplot(LitterDataSpreadNeedles) +
  geom_point(aes(x = collectDate, y = Needles)) +
  facet_wrap(vars(nlcdClass), nrow = 3)
print(NeedleMassFacetPlot)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think the faceted plot is more effective, because when all of the data points are overlapping with eachother it looks clutterd and it is difficult to view the differences between the three different colors. When there are facets it is easier to view the differences between the three different classes. 